BERT Model Compression and Weight Sharing

### 2019-05-13 paper/paper.tex
Rewriting literature review and inserting matrix spectrum picture

DNNs have become extremely large in size with growing complexities. A BERT
base model has 110M parameters and takes 438 MB disk space to store. This
makes it very hard to deploy those models on mobile devices, which in most
cases have limited storage and RAM and low computational resources. As a
result, we are not able to store multiple models for different tasks on mobile
devices. Allocating hundreds of megabytes to to those models is also not
realistic for mobile apps such as Gmail or Facebook. And due to increasing
privacy concerns, users prefer models generating predictions based on their
input to be running locally instead of sending information to the cloud. Last
but not least, evaluating enormous models takes too much time on mobile
devices and can drain batteries significantly.

The challenge that we are facing is therefore how to compress DNNs in size
while keeping their predictive performance competitive with full model.

@@

We are interested in compressing the language representation model BERT
\citep{devlin2018bert}.  It is the state-of-art result in many natural
language processing tasks  such as natural language inference, paraphrasing
and named entity recognition.

\citet{cheng2017survey} provides a comprehensive overview of the state-of-art
methods to compress DNNs. Of the methods summarized by the paper, the ones
that are of interest to our work are weight-pruning and quantization
\citep{han2015deep}, and low-rank factorization \citep{denton2014exploiting}.
Knowledge distillation, first proposed by \citep{bucilua2006model} as
``knowledge transfer'', was recently popularized by
\citep{hinton2015distilling}. It differs from the previously mentioned methods
since it compresses an existing model by replacing it with a simpler model
that is trained on its outputs.

DeepTwist, a end-to-end training framework, allows us to incorporate
compression procedures into training \citep{lee2018deeptwist}. Under this
framework, the model parameters are occasionally distorted during training
time instead of only at the end. This framework has the benefit of simplified
training procedure and faster convergence and reduced requirement for
hyperparameter tuning. Experiment on MNIST dataset with DeepTwist shows that
LeNet-5 and LeNet-300-100 models are able to retain predictive accuracy under
high compression rate.

@@

% \begin{table}[tbh]
%         \caption{Results for DeepTwist}
%         \label{tab:deeptwist_results}
%         \centering
%         \vspace{1em}
%         \begin{tabular}{lccccc}
% \toprule
% {} &               type &  param &  val loss &   val acc \\
% {} & 
% \hypertarget{tabcol:1}{(1)} & \hypertarget{tabcol:2}{(2)} & \hypertarget{tabcol:3}{(3)} & \hypertarget{tabcol:4}{(4)} \\ \midrule
% 0  &  diff prune global &   $90$ &   $0.211$ &   $0.912$ \\
% 1  &              prune &   $20$ &   $0.291$ &   $0.886$ \\
% 2  &              prune &   $80$ &   $0.532$ &   $0.767$ \\
% 3  &         diff prune &   $80$ &   $0.222$ &   $0.916$ \\
% 4  &  diff prune global &   $99$ &       --- &       --- \\
% 5  &  diff prune global &   $90$ &   $0.213$ &   $0.916$ \\
% 6  &         diff prune &   $40$ &   $0.203$ &   $0.909$ \\
% 7  &                svd &   $50$ &   $0.490$ &   $0.810$ \\
% 8  &                svd &  $200$ &   $0.459$ &   $0.845$ \\
% 9  &              prune &   $40$ &   $0.424$ &   $0.845$ \\
% 10 &              prune &   $60$ &   $0.425$ &   $0.815$ \\
% 11 &  diff prune global &   $99$ &   $0.230$ &   $0.901$ \\
% 12 &  diff prune global &   $99$ &   $0.222$ &   $0.905$ \\
% 13 &         diff prune &   $99$ &   $0.222$ &   $0.913$ \\
% 14 &         diff prune &   $95$ &   $0.221$ &   $0.916$ \\
% 15 &         diff prune &   $90$ &   $0.216$ &   $0.916$ \\
% 16 &         diff prune &   $60$ &   $0.209$ &   $0.920$ \\
% 17 &                svd &  $100$ &   $0.515$ &   $0.823$ \\
% 18 &                svd &  $150$ &   $0.456$ &   $0.828$ \\
% 19 &         diff prune &   $80$ &   $2.533$ &   $0.491$ \\
% 20 &         diff prune &   $90$ &       --- &       --- \\
% 21 &  diff prune global &    $0$ &   $0.738$ &   $0.509$ \\
% 22 &         diff prune &    $0$ &   $0.692$ &   $0.522$ \\
% 23 &         diff prune &    $0$ &   $1.854$ &   $0.491$ \\
% \bottomrule
% \end{tabular}

@@

Our non-difference distortions are more sensitive to the compression parameter
(whether prune rate or SVD rank). We find it difficult to achieve performance
comparable to the uncompressed BERT with a significantly compressed model. Among
the different distortion methods, we favor pruning for its computational
simplicity, and see no evidence that SVD achieves better compression.

>>>>

DNNs have become extremely large in size with growing complexities. A good
example is BERT \citep{devlin2018bert}, which represents a growing
literature in natural language processing in favor of pretraining 
\citep{peters2018deep,radford2018improving,radford2019language}, which
uses large models to learn from a large corpus of text in an unsupervised
manner. The goal of such pretraining is creating contextualized
representation of text in $\R^n$ which preserve certain structure of
language. For different tasks, the researcher then \emph{fine-tunes} the
pretrained model to the task, which may reach good performance since the
pretrained representations are well-behaved representations of language
strucutre. A downside of this framework is that each fine-tuned task
requires a unique set of weights, which, when coupled with the raw size of
these models, are cumbersome objects especially in a low hardware resource
environment.

The smaller \texttt{BERT-base-uncased} model has 110M parameters and takes
438 MB disk space to store. This
makes it very hard to deploy those models on mobile devices, which in most
cases have limited storage and RAM and low computational resources. As a
result, we are not able to store multiple models for different tasks on mobile
devices. Allocating hundreds of megabytes to to those models is also not
realistic for mobile apps such as Gmail or Facebook. And due to increasing
privacy concerns, users prefer models generating predictions based on their
input to be running locally instead of sending information to the cloud. Last
but not least, evaluating enormous models takes too much time on mobile
devices and can drain batteries significantly.

The challenge that we are facing is therefore how to compress DNNs in size
while keeping their predictive performance competitive with full model. We are interested in compressing the language representation model BERT.  It is the state-of-art result in many natural
language processing tasks  such as natural language inference, paraphrasing
and named entity recognition.

@@

There is a rich literature on compressing deep learning models, a selection
of which we review in this section. For a comprehensive review, see  
\citep{cheng2017survey}. The early work in compression is usually in a
multi-stage framework, where model is trained, compressed, and then
the compressed model is retrained for good performance.

\citet{han2015deep} applies compression techniques to computer vision,
where convolutional neural networks are popular and are often
overparameterized \citep{yang2019fsnet},
where the network is gradually \emph{pruned} and retrained in the first
stage, where small
weights are set
to zero and gradient connections severed. In the second stage, the weights
that survived pruning are clustered, and each cluster behaves like a single
weight, with identical gradients. In the last stage, the model is Huffman
encoded. Numerical experiments by \citet{han2015deep} show that the
compression does not seem to influence performance. Moreover, the benefits
of various compression techniques appear to stack without much loss in
compression benefit. Moreover, convolutional network sizes are dominated by
the convolution operations in
lower layers \citep{denton2014exploiting}. \citet{denton2014exploiting}
uses matrix factorization techniques on the matrices in convolutional
layers, which has low effective rank.

Pruning, quantization, and rank reduction are operations applied to a
particular network's weights without altering the network architecture. 
A different framework, knowledge distillation, first proposed by 
\citep{bucilua2006model} as
``knowledge transfer,'' was recently popularized by
\citep{hinton2015distilling}. In knowledge distillation, a simple network,
called the \emph{student network}, is not trained on the actual data but on
the output of the large network one wishes to compress, called the 
\emph{teacher network}. This procedure can even be extended by inserting
various
intermediate-sized networks between the teacher and student, intuitively
termed
\emph{teacher-assistant networks} \citep{mirzadeh2019improved}. We dicuss
knowledge distillation in detail in \Cref{sec:training}.

Compression techniques like pruning, quantization, and rank reduction can
also be incorporated into an end-to-end framework, known as 
\emph{DeepTwist}  \citep{lee2018deeptwist}, where training needs not be
interrupted in some multi-stage process. Under this
framework, the model parameters are occasionally distorted during training
time instead of only at the end. This framework has the benefit of simplified
training procedure and faster convergence and reduced requirement for
hyperparameter tuning. Experiment on MNIST dataset with DeepTwist shows that
LeNet-5 and LeNet-300-100 models are able to retain predictive accuracy under
high compression rate.

@@

% \begin{table}[tbh]
%         \caption{Results for DeepTwist}
%         \label{tab:deeptwist_results}
%         \centering
%         \vspace{1em}
%         \begin{tabular}{lccccc}
% \toprule
% {} &               type &  param &  val loss &   val acc \\
% {} & 
% \hypertarget{tabcol:1}{(1)} & \hypertarget{tabcol:2}{(2)} & \hypertarget{tabcol:3}{(3)} & \hypertarget{tabcol:4}{(4)} \\ \midrule
% 0  &  diff prune global &   $90$ &   $0.211$ &   $0.912$ \\
% 1  &              prune &   $20$ &   $0.291$ &   $0.886$ \\
% 2  &              prune &   $80$ &   $0.532$ &   $0.767$ \\
% 3  &         diff prune &   $80$ &   $0.222$ &   $0.916$ \\
% 4  &  diff prune global &   $99$ &       --- &       --- \\
% 5  &  diff prune global &   $90$ &   $0.213$ &   $0.916$ \\
% 6  &         diff prune &   $40$ &   $0.203$ &   $0.909$ \\
% 7  &                rank reduction &   $50$ &   $0.490$ &   $0.810$ \\
% 8  &                rank reduction &  $200$ &   $0.459$ &   $0.845$ \\
% 9  &              prune &   $40$ &   $0.424$ &   $0.845$ \\
% 10 &              prune &   $60$ &   $0.425$ &   $0.815$ \\
% 11 &  diff prune global &   $99$ &   $0.230$ &   $0.901$ \\
% 12 &  diff prune global &   $99$ &   $0.222$ &   $0.905$ \\
% 13 &         diff prune &   $99$ &   $0.222$ &   $0.913$ \\
% 14 &         diff prune &   $95$ &   $0.221$ &   $0.916$ \\
% 15 &         diff prune &   $90$ &   $0.216$ &   $0.916$ \\
% 16 &         diff prune &   $60$ &   $0.209$ &   $0.920$ \\
% 17 &                rank reduction &  $100$ &   $0.515$ &   $0.823$ \\
% 18 &                rank reduction &  $150$ &   $0.456$ &   $0.828$ \\
% 19 &         diff prune &   $80$ &   $2.533$ &   $0.491$ \\
% 20 &         diff prune &   $90$ &       --- &       --- \\
% 21 &  diff prune global &    $0$ &   $0.738$ &   $0.509$ \\
% 22 &         diff prune &    $0$ &   $0.692$ &   $0.522$ \\
% 23 &         diff prune &    $0$ &   $1.854$ &   $0.491$ \\
% \bottomrule
% \end{tabular}

@@

Our non-difference distortions are more sensitive to the compression parameter
(whether prune rate or rank reduction rank). We find it difficult to achieve performance
comparable to the uncompressed BERT with a significantly compressed model. Among
the different distortion methods, we favor pruning for its computational
simplicity, and see no evidence that rank reduction achieves better compression.

Remarkably, the fact that pruning and rank reduction incurs significant
losses to the
model seems to show that popular compression techniques that enjoy success
in convolutional neural networks fail to generalize good performance to
transformer-based language networks like BERT. We present some preliminary
evidence in \Cref{fig:singular-values}, where we plot the singular values
in one of the matrices in BERT. The effective rank of the matrix seems
quite high, as the singular values are all sizable, decaying at a rate too
slow for rank reduction techniques to be promising. This suggests that
matrices in BERT are in fact effectively dense, unlike CNNs, where
good compression relies on the effective sparsity of the matrices. 
\begin{figure}[tb]
    \centering
    \includegraphics[width=0.5\textwidth]
    {../figs/svd-singular-vals-distribution.png}
    \caption{Distribution of singular values in one of BERT layers}
    \label{fig:singular-values}
\end{figure}
